---
title: "Week 14: Data Science and Music 🎵🎸👩‍🎤"
author: "Eren Bilen, Assistant Professor of Data Analytics"
institute: "Data 101: Data Science for Humanities and Social Sciences <br> Dickinson College"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: insert-logo.html
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(reticulate)

# point reticulate to the exact Python you want
py <- "/Users/eren/anaconda3/envs/r-reticulate/bin/python"
Sys.setenv(RETICULATE_PYTHON = py)
use_python(py, required = TRUE)

# make knitr use the same Python for python-engine chunks
knitr::opts_chunk$set(engine.path = list(python = py))

#use_python("C:/Users/Eren/AppData/Local/r-miniconda/envs/r-reticulate/python.exe", required = TRUE)

# install once if needed (safe if already installed)
#try(py_install(c("pandas", "matplotlib", "seaborn"), pip = TRUE), silent = TRUE)

#py_config()   # sanity check in knit log
```


# Music... what is it?

--

- A collection of sounds to produce harmony, melodies, and emotions.

--

- Sound is quantifiable. How?
  - Every sound is a vibration in a certain frequency.
    - Frequency is measured in Hertz (Hz). Humans can hear between 20 to 20,000 Hz. Dogs 40 to 60,000 Hz. 
  - Higher the sound/pitch, higher frequency. We can measure this.
  - Notes are sorted by pitch: C natural scale: C D E F G A B
  - Rythm and tempo are also measurable. Eg, a song might be 120 bpm (beats per minute)
  
---

# How does it work in Python?

- Let's read in real music into python. 
- We need to import songs. Watch out for copyright! 
- GTZAN dataset contains 100 free songs from 10 genres. Songs were collected between 2000-01. Publicly available [here](https://www.kaggle.com/datasets/andradaolteanu/gtzan-dataset-music-genre-classification).

```{python}
import librosa # very popular package to analyze audio files
# Read a reggae song
audio = "reggae.00010.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "reggae.00010.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

---

```{python}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('Redemption Song by Bob Marley')
plt.show()
```

---

```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---

# Song 2

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "blues.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "blues.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A standard Blues song')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---

# Song 3

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "classical.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "classical.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A classical piece')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---


# Song 4

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "country.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "country.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A country song')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---

# Song 5

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "disco.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "disco.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('I Will Survive by Gloria Gaynor')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---

# Song 6

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "hiphop.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "hiphop.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A hiphop song')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```


---

# Song 7

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "jazz.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "jazz.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A jazz song')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---

# Song 8

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "metal.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "metal.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A metal song')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```


---

# Song 9

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "pop.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "pop.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('A pop song')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```


---

# Song 10

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "reggae.00001.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "reggae.00001.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('No Woman No Cry by Bob Marley')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```


---

# Song 11

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
# Read the song
audio = "rock.00026.wav" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "rock.00026.wav"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('Radio Ga Ga by Queen')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```



---

# Bonus: A song covered by me!

```{python, echo=FALSE}
import librosa # very popular package to analyze audio files
import warnings
warnings.filterwarnings("ignore", category=UserWarning, module="librosa")
warnings.filterwarnings("ignore", category=FutureWarning, module="librosa")
warnings.filterwarnings("ignore", message="PySoundFile failed")
# Read the song
audio = "happyjusttodance.m4a" # waveform audio file
data, sr = librosa.load(audio) # data has freqs, sr is sampling rate
```

```{r, echo=FALSE}
library(htmltools)
audio_file <- "happyjusttodance.m4a"   # put the file in the same folder
htmltools::tags$audio(
  src = audio_file,
  type = "audio/wav",
  controls = NA
)
```

```{python, echo=FALSE}
# Wave form of the audio
import matplotlib.pyplot as plt
plt.figure(figsize=(7,4))
librosa.display.waveshow(data,color="#2B4F72", alpha = 0.5);
plt.ylabel('amplitude')
plt.title('Happy Just to Dance With You')
plt.show()
```

---
```{python}
# Spectrogram of the audio
stft=librosa.stft(data)
stft_db=librosa.amplitude_to_db(abs(stft))
plt.figure(figsize=(7,4))
librosa.display.specshow(stft_db,sr=sr,x_axis='time',y_axis='hz');
plt.colorbar();
```

---

<div style="font-size:10px; line-height:1">
<style>
.slide-specific pre code {
  font-size: 10px !important;
}
</style>

<div class="slide-specific">

```{python,echo=FALSE}
import warnings
import librosa
warnings.filterwarnings("ignore", message="librosa.beat.tempo")
```

```{python}
import librosa
import numpy as np
import pandas as pd

def extract_features(filename):
    audio, sr = librosa.load(filename, duration=30)

    # Important: use keyword arguments so it works with new librosa versions
    tempo = librosa.beat.tempo(y=audio, sr=sr)[0]
    chroma = np.mean(librosa.feature.chroma_stft(y=audio, sr=sr))
    rmse = np.mean(librosa.feature.rms(y=audio))
    centroid = np.mean(librosa.feature.spectral_centroid(y=audio, sr=sr))
    bandwidth = np.mean(librosa.feature.spectral_bandwidth(y=audio, sr=sr))
    rolloff = np.mean(librosa.feature.spectral_rolloff(y=audio, sr=sr))
    zcr = np.mean(librosa.feature.zero_crossing_rate(y=audio))

    mfcc_values = librosa.feature.mfcc(y=audio, sr=sr, n_mfcc=5)
    mfcc1 = np.mean(mfcc_values[0])
    mfcc2 = np.mean(mfcc_values[1])
    mfcc3 = np.mean(mfcc_values[2])
    mfcc4 = np.mean(mfcc_values[3])
    mfcc5 = np.mean(mfcc_values[4])

    return pd.Series({
        "tempo": tempo,
        "chroma": chroma,
        "rmse": rmse,
        "centroid": centroid,
        "bandwidth": bandwidth,
        "rolloff": rolloff,
        "zero_crossing_rate": zcr,
        "mfcc1": mfcc1,
        "mfcc2": mfcc2,
        "mfcc3": mfcc3,
        "mfcc4": mfcc4,
        "mfcc5": mfcc5
    })

# ---------------------------------------------------------------
# Extract features for the songs
blues = extract_features("/Users/eren/Downloads/Data/genres_original/blues/blues.00001.wav")
classical = extract_features("/Users/eren/Downloads/Data/genres_original/classical/classical.00001.wav")
jazz = extract_features("/Users/eren/Downloads/Data/genres_original/jazz/jazz.00001.wav")
country = extract_features("/Users/eren/Downloads/Data/genres_original/country/country.00001.wav")
disco = extract_features("/Users/eren/Downloads/Data/genres_original/disco/disco.00001.wav")
rock = extract_features("/Users/eren/Downloads/Data/genres_original/rock/rock.00026.wav")
recording = extract_features("/Users/eren/Downloads/happyjusttodance.m4a")

songs_df = pd.DataFrame({
    "blues": blues,
    "classical": classical,
    "jazz": jazz,
    "country": country,
    "disco": disco,
    "rock": rock,
    "recording": recording
})
```

---

```{python,echo=FALSE}
import matplotlib.pyplot as plt
import seaborn as sns

df = songs_df.copy()

# reshape
df_reset = df.reset_index().rename(columns={'index': 'feature'})
df_long = df_reset.melt(id_vars='feature', var_name='genre', value_name='value')

features = [
    "tempo",
    "chroma",
    "rmse",
    "centroid",
    "bandwidth",
    "rolloff",
    "zero_crossing_rate"
]

sns.set_style("white")

fig, axes = plt.subplots(nrows=4, ncols=2, figsize=(6, 9))
axes = axes.flatten()

# Strong, saturated palette
palette = sns.color_palette("gnuplot2")   # bold, high contrast

for i, feat in enumerate(features):
    subset = df_long[df_long["feature"] == feat]

    sns.barplot(
        data=subset,
        x="genre", y="value",
        ax=axes[i],
        palette=palette
    )

    axes[i].set_title(feat, fontsize=15)
    axes[i].tick_params(axis='x', rotation=45)
    axes[i].set_xlabel("")
    axes[i].set_ylabel("")
    axes[i].grid(False)
    axes[i].set_facecolor("white")

fig.delaxes(axes[-1])
plt.tight_layout()
plt.show()


```

---

# Other interesting applications:

- With text data, we said: Every author leaves a measurable statistical “signature.” (stylometry)
  - Same is true with music!
  - Note level analysis: number of notes, chords, changes in keys...
- Genres over time? Has pop been the same over the last century?
- Machine learning and AI: Google's Bach bot
- Music and culture: The development of western music and eastern music
- Music lyrics (text analysis)

🤑 Music recommendation algorithms, song recognizer apps, eg, Shazam


